Using Heavy-Tailed Self-Regularization Theory to Characterize Correlations in and Perform Diagnostics on State-of-the-art Overparameterized Models

### 2021-02-18 submission/TOPML_Workshop/paper.tex
paper on four pages, lots of fat removed

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{A Differential Geometry Perspective on Orthogonal Recurrent Models}

@@

\twocolumn[
\icmltitle{XXX TITLE}

@@

\begin{abstract}
XXX.  ABSTRACT.  THE NOMINAL CONTRIBUTION IS TO APPLY HT-SR THEORY AND WEIGHTWATCHER TO DO AN ANALYSIS OF DATA ANS IDENTIFY SIMPSONS LIKE PARADOXES.
\end{abstract}

@@

XXX.  INTRO AND OVERVIEW, BASICALLY THE CORE DUMP CHARLES DID ON THE PHONE.

\michael{Charles, put in a core dump of the intro that you said.  Maybe half a column to one column, in this two column format.}

@@

XXX.  SUMMARY OF STAT MECH AND HTSR THEORY FOR DNNS.

\cite{MM17_TR} is Charles: Rethinking generalization requires revisiting old ideas: statistical mechanics approaches and complex learning behavior''

\cite{MM18_TR} is Charles: ``Implicit Self-Regularization in Deep Neural Networks: Evidence from Random Matrix Theory and Implications for Learning''

\cite{MM19_HTSR_ICML} is Charles: ``Traditional and Heavy-Tailed Self Regularization in Neural Network Models''

\cite{MM20_SDM} is Charles: ``Heavy-Tailed {U}niversality Predicts Trends in Test Accuracies for Very Large Pre-Trained Deep Neural Networks''

\cite{MM19_KDD} is Charles: ``Statistical Mechanics Methods for Discovering Knowledge from Modern Production Quality Neural Networks''

XXX.  ABSTRACT FROM \cite{MM18_TR}:
Random Matrix Theory (RMT) is applied to analyze the weight matrices of Deep Neural Networks (DNNs), including both production quality, pre-trained models such as AlexNet and Inception, and smaller models trained from scratch, such as LeNet5 and a miniature-AlexNet.
Empirical and theoretical results clearly indicate that the DNN training process itself implicitly implements a form of \emph{Self-Regularization}, implicitly sculpting a more regularized energy or penalty landscape.
In particular, the empirical spectral density (ESD) of DNN layer matrices displays signatures of traditionally-regularized statistical models, even in the absence of exogenously specifying traditional forms of explicit regularization, such as Dropout or Weight Norm constraints.
Building on relatively recent results in RMT, most notably its extension to Universality classes of Heavy-Tailed matrices, and applying them to these empirical results,
we develop a theory to identify \emph{5+1 Phases of Training}, corresponding to increasing amounts of \emph{Implicit Self-Regularization}.
These phases can be observed during the training process as well as in the final learned DNNs.
For smaller and/or older DNNs, this Implicit Self-Regularization is like traditional Tikhonov regularization, in that there is a ``size scale'' separating signal from noise.
For state-of-the-art DNNs, however, we identify a novel form of \emph{Heavy-Tailed Self-Regularization}, similar to the self-organization seen in the statistical physics of disordered systems (such as classical models of actual neural activity).
This results from correlations arising at all size scales, which for DNNs arises implicitly due to the training process itself.
This implicit Self-Regularization can depend strongly on the many knobs of the training process.
In particular, by exploiting the generalization gap phenomena,
we demonstrate that we can cause a small model to exhibit all 5+1 phases of training simply by changing the batch size.
This demonstrates that---all else being equal---DNN optimization with larger batch sizes leads to less-well implicitly-regularized models, and it provides an explanation for the generalization gap phenomena.
Our results suggest that large, well-trained DNN architectures should exhibit Heavy-Tailed Self-Regularization, and we discuss the theoretical and practical implications of~this.

\subsection{Overall approach}
\label{sxn:approach}

XXX.  THE FOLLOWING IS TROM THE PREDICTING TRENDS PAPER.

Consider the objective/optimization function (parameterized by $\mathbf{W}_{l}$s and $\mathbf{b}_{l}$s) for a DNN with $L$ layers, and weight matrices $\mathbf{W}_{l}$ and bias vectors $\mathbf{b}_{l}$, as 
the minimization of a general loss function $\mathcal{L}$ over the training data instances and labels, $\{\mathbf{x}_{i},y_{i}\}\in\mathcal{D}$.
For a typical supervised classification problem, the goal of training is to construct (or learn) $\mathbf{W}_{l}$ and $\mathbf{b}_{l}$ that capture correlations in the data, in the sense of solving
\begin{equation}
%%%\underset{\mathbf{W}_{l},\mathbf{b}_{L}}{\text{argmin}}\;\sum_{i=1}^{N}\left(\mathcal{L}(E_{DNN}(\mathbf{x}_{i})-y_{i})\right)  ,
\underset{\mathbf{W}_{l},\mathbf{b}_{L}}{\text{argmin}}\;\sum_{i=1}^{N} \mathcal{L}(E_{DNN}(\mathbf{x}_{i}),y_{i})   ,
\end{equation}
where the loss function $\mathcal{L}(\cdot,\cdot)$ can take on a myriad of forms, and where the energy (or optimization) landscape function
\begin{equation}
%E_{DNN}=h_{L}(\mathbf{W}_{L}\times h_{L-1}(\mathbf{W}_{L-1}\times h_{L-2}(\cdots)+\mathbf{b}_{L-1})+\mathbf{b}_{L})
E_{DNN} = f(\mathbf{x}_{i} ; \mathbf{W}_{1},\ldots,\mathbf{W}_{L},\mathbf{b}_{1},\ldots,\mathbf{b}_{L})
\label{eqn:dnn_energy}
\end{equation}
depends parametrically on the weights and biases.
For a trained model, the form of the function $E_{DNN}$ does not explicitly depend on the data (but it does explicitly depend on the weights and biases).
The function $E_{DNN}$ maps data instance vectors ($\mathbf{x}_i$ values) to predictions ($y_{i}$ labels), and thus the output of this function does depend on the data.
Therefore, one can analyze the form of $E_{DNN}$ in the absence of any training or test~data.

Test accuracies have been reported online for publicly-available pretrained pyTorch models~\cite{osmr}.
These models have been trained and evaluated on labeled data $\{\mathbf{x}_{i},y_{i}\}\in\mathcal{D}$, using standard techniques.  
We do not have access to this data, and we have not trained any of the models ourselves. 
Our methodological approach is thus similar to a statistical meta-analysis, common in biomedical research, but uncommon in ML.
%
Computations were performed with the publicly-available \texttt{WeightWatcher} tool (version 0.2.7)~\cite{weightwatcher_package}.

\paragraph{Metrics for DNN Weight Matrices.}

Our approach involves analyzing individual DNN weight matrices, for (depending on the architecture) fully-connected and/or convolutional layers.
Each DNN layer contains one or more layer 2D  $N_{l}\times M_{l}$ weight matrices, $\mathbf{W}_{l}$, or pre-activation maps, $\mathbf{W}_{i,l}$, e.g., extracted from 2D Convolutional layers, where $N > M$.
See the Supplementary Information
%~\ref{sxn:appendix} 
for details.
(We may drop the $i$ and/or $i,l$ subscripts below.)
The best performing quality metrics depend on the norms and/or spectral properties of each weight matrix,
$\mathbf{W}$, and/or, equivalently, it's empirical correlation matrix, $\mathbf{X}=\mathbf{W}^{T}\mathbf{W}$.
To evaluate the quality of state-of-the-art DNNs, 
we consider the following metrics:
\begin{eqnarray}
& & \text{Frobenius Norm: $\Vert\mathbf{W}\Vert^{2}_{F}=\Vert\mathbf{X}\Vert_{F}=\sum\nolimits_{i=1}^{M} \lambda_{i}$ } \\
& & \text{Spectral Norm: $\Vert\mathbf{W}\Vert_{\infty}^{2}=\Vert\mathbf{X}\Vert_{\infty}=\lambda_{max}$ } \\
& & \text{Weighted Alpha: $\hat{\alpha}=\alpha\log\lambda_{max}$ } \\
& & \text{$\alpha$-Norm (or $\alpha$-Shatten Norm): $\Vert\mathbf{W}\Vert^{2\alpha}_{2\alpha}=\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}=\sum\nolimits_{i=1}^{M}\lambda_{i}^{\alpha}$. }
\end{eqnarray}
%To diagnose problems, 
To perform diagnostics on potentially-problematic DNNs,
we will decompose $\hat{\alpha}$ into its two components, $\alpha$ and $\lambda_{max}$.
Here, $\lambda_{i}$ is the $i^{th}$ eigenvalue of the $\mathbf{X}$, $\lambda_{max}$ is the maximum eigenvalue, and $\alpha$ is the fitted PL exponent. 
These eigenvalues are squares of the singular values $\sigma_{i}$ of $\mathbf{W}$, $\lambda_{i}=\sigma^{2}_{i}$.
All four metrics can be computed easily from DNN weight matrices.
The first two metrics are well-known in ML.
The last two metrics deserve special mention, as they depend on an empirical parameter $\alpha$ that is the PL exponent that arises in the recently-developed Heavy Tailed Self Regularization (HT-SR) Theory~\cite{MM18_TR, MM19_HTSR_ICML, MM20_SDM}.

\paragraph{Overview of Heavy-Tailed Self-Regularization.}

In the HT-SR Theory, one analyzes the eigenvalue spectrum, i.e., the Empirical Spectral Density (ESD), of the associated correlation matrices~\cite{MM18_TR,MM19_HTSR_ICML,MM20_SDM}.
From this, one characterizes the amount and form of correlation, and therefore implicit self-regularizartion, present in the DNN's weight matrices.
For each layer weight matrix $\mathbf{W}$, of size $N \times M$, construct the associated $M\times M$ (uncentered) correlation matrix $\mathbf{X}$. 
Dropping the $L$ and $l,i$ indices, one~has
$$
\mathbf{X} = \frac{1}{N}\mathbf{W}^{T}\mathbf{W}.
$$
If we compute the eigenvalue spectrum of $\mathbf{X}$, i.e., $\lambda_i$ such that
$  % $$
\mathbf{X}\mathbf{v}_{i}=\lambda_{i}\mathbf{v}_{i} , 
$  % $$
then the ESD of eigenvalues, $\rho(\lambda)$, is just a histogram of the eigenvalues, formally written as
%\begin{equation}
$\rho(\lambda)=\sum\nolimits_{i=1}^{M}\delta(\lambda-\lambda_{i})  .$
%\label{eqn:eigenval_hist}
%\end{equation}
Using HT-SR Theory, one characterizes the correlations in a weight matrix by examining its ESD, $\rho(\lambda)$.
It can be well-fit to a truncated PL distribution, given~as
\begin{equation}
\rho(\lambda)\sim\lambda^{-\alpha}  ,
\label{eqn:eigenval_pl}
\end{equation}
which is (at least) valid within a bounded range of eigenvalues $\lambda\in[\lambda^{min},\lambda^{max}]$.

The original work on HT-SR Theory considered a small number of NNs, including AlexNet and InceptionV3. 
It showed that for nearly every $\mathbf{W}$, the (bulk and tail) of the ESDs can be fit to a truncated PL, and that PL exponents $\alpha$ nearly all lie within the range $\alpha\in(1.5,5)$~\cite{MM18_TR,MM19_HTSR_ICML,MM20_SDM}.
%
%%As for the mechanism responsible for these properties, statistical physics offers several possibilities~\cite{SornetteBook,nishimori01}, e.g., self organized criticality~\cite{SOC87,SOCat25yrs} or multiplicative noise in the stochastic optimization algorithms used to train these models~\cite{HodMah20A_TR,SorCon97}.
%%Alternatively, related techniques have been used to analyze correlations and information propogation in actual spiking neurons~\cite{SYYRP11,YKYP14}.
%%Our meta-analysis does not require knowledge of mechanisms; and it is not even clear that one mechanism is responsible for every case.
%
Crucially, HT-SR Theory predicts that smaller values of $\alpha$ should correspond to models with better correlation over multiple size scales and thus to better models.
The notion of ``size scale'' is well-defined in physical systems, to which this style of analysis is usually applied, but it is less well-defined in CV and NLP applications.
Informally, it would correspond to pixel groups that are at a greater distance in some metric, or between sentence parts that are at a greater distance in text.
Relatedly, previous work observed that smaller exponents $\alpha$ correspond to more implicit self-regularization and better generalization, and that we expect a linear correlation between $\hat{\alpha}$ and model quality~\cite{MM18_TR,MM19_HTSR_ICML,MM20_SDM}.

\paragraph{DNN Empirical Quality Metrics.}

For norm-based metrics, we use the average of the log norm, to the appropriate power.
Informally, this amounts to assuming that the layer weight matrices are statistically independent, in which case we can estimate the model complexity $\mathcal{C}$, or test accuracy, with a standard Product Norm (which resembles a data dependent VC complexity),
\begin{equation}
\mathcal{C}\sim\Vert\mathbf{W}_{1}\Vert\times\Vert\mathbf{W}_{2}\Vert \times \cdots \times \Vert\mathbf{W}_{L}\Vert ,
\end{equation}
where $\Vert\cdot\Vert$ is a matrix norm.   
The log complexity,
\begin{equation}
\label{eqn:eqn:sum_log_norm}
\log\mathcal{C} \sim \log\Vert\mathbf{W}_{1}\Vert+\log\Vert\mathbf{W}_{2}\Vert + \cdots + \log\Vert\mathbf{W}_{L}\Vert = \sum\nolimits_l \log\Vert\mathbf{W}_{l}\Vert ,
\end{equation}
 takes the form of an average Log Norm.
For the Frobenius Norm metric and Spectral Norm metric, we can use Eqn.~(\ref{eqn:eqn:sum_log_norm}) directly (since, when taking $\log\Vert\mathbf{W}_{l}\Vert_{F}^{2}$, the $2$ comes down and out of the sum, and thus ignoring it only changes the metric by a constant factor).

The Weighted Alpha metric is an average of $\alpha_l$ over all layers $l \in \{1,\ldots,l\}$, weighted by the size, or scale, or each matrix,
\begin{equation}
\hat{\alpha} = \dfrac{1}{L}\sum_l \alpha_l\log\lambda_{max,l}\approx\langle\log\Vert\mathbf{X}\Vert_{\alpha}^{\alpha}\rangle    ,
\end{equation}
where $L$ is the total number of layer weight matrices.
The Weighted Alpha metric was introduced previously~\cite{MM20_SDM}, where it was shown to correlate well with trends in reported test accuracies of pretrained DNNs, albeit on a much smaller and more limited set of models than we consider here.

Based on this, in this paper, we introduce and evaluate the $\alpha$-Shatten Norm metric,
\begin{equation}
\label{eqn:sum_log_alpha_norm_alpha}
\sum\nolimits_l \log \Vert\mathbf{X}_l\Vert_{\alpha_l}^{\alpha_l} 
=
\sum\nolimits_l \alpha_l \log \Vert\mathbf{X}_l\Vert_{\alpha_l} .
\end{equation}
For the $\alpha$-Shatten Norm metric, $\alpha_l$ varies from layer to layer, and so in Eqn.~(\ref{eqn:sum_log_alpha_norm_alpha}) it cannot be taken out of the sum.
For small $\alpha$, the Weighted Alpha metric approximates the Log $\alpha$-Shatten norm, as can be shown with a statistical mechanics and random matrix theory derivation;
%% \cite{MM20_unpub_work}; 
and the Weighted Alpha and $\alpha$-Shatten norm metrics often behave like an improved, weighted average Log Spectral Norm.

Finally, although it does less well for predicting trends in state-of-the-art model series, e.g., as depth changes, the average value of $\alpha$, i.e., 
\begin{equation}
\bar{\alpha} = \dfrac{1}{L}\sum_l \alpha_l = \langle\alpha\rangle    ,
\label{eqn:alpha_bar}
\end{equation}
can be used to perform model diagnostics, to identify problems that cannot be detected by examining training/test accuracies, and to discriminate poorly-trained models from well-trained~models.

One determines $\alpha$ for a given layer by fitting the ESD of that layer's weight matrix to a truncated PL, using the commonly accepted Maximum Likelihood method~\cite{CSN09_powerlaw,ABP14}.
%%This method works very well for exponents between $\alpha\in(2,4)$; and it is adequate, although imprecise, for smaller and especially larger $\alpha$~\cite{newman2005_zipf}. 
%
Operationally, $\alpha$ is determined by using the \texttt{WeightWatcher} tool~\cite{weightwatcher_package} to fit the histogram of eigenvalues, $\rho(\lambda)$, to a truncated PL, 
%%\begin{equation}
%%\rho(\lambda)\sim\lambda^{\alpha},\;\;\lambda\in[\lambda_{min},\lambda_{max}] ,
%%\end{equation}
%%where $\lambda_{max}$ is the largest eigenvalue of $\mathbf{X}=\mathbf{W}^{T}\mathbf{W}$, and 
%%where $\lambda_{min}$ is selected automatically to yield the best (in the sense of minimizing the K-S distance) PL fit.
%%Each of these quantities is defined for a given layer $\mathbf{W}$ matrix.
%%See Figure~\ref{fig:ww} for an illustration.

%\begin{figure}[t]
%    \centering
%    %\includegraphics[width=15.0cm]{img/WeightWatcher_v2}
%    \includegraphics[width=15.0cm]{img/WeightWatcher_v3}
%    \caption{Schematic of analyzing DNN layer weight matrices $\mathbf{W}$.  
%             Given an individual layer weight matrix $\mathbf{W}$, from either a fully-connected layer or a convolutional layer, perform a Singular Value Decomposition (SVD) to obtain $\mathbf{W} = \mathbf{U} \mathbf{\Sigma} \mathbf{V}^{T}$, and examine the histogram of eigenvalues of $\mathbf{W}^{T}\mathbf{W}$.
%             Norm-based metrics and PL-based metrics (that depend on fitting the histogram of eigenvalues to a truncated PL) can be used to compare models.
%             For example, one can analyze one layer of a pre-trained model, compare multiple layers of a pre-trained model, make comparisons across model architectures, monitor neural network properties during training, etc. 
%}
%    \label{fig:ww}
%\end{figure}

To avoid confusion, let us clarify the relationship between $\alpha$ and $\hat{\alpha}$.  
We fit the ESD of the correlation matrix $\mathbf{X}$ to a truncated PL, parameterized by 2 values: the PL exponent $\alpha$, and the maximum eigenvalue $\lambda_{max}$.
The PL exponent $\alpha$ measures the amount of correlation in a DNN layer weight matrix $\mathbf{W}$. 
It is valid for $\lambda\le\lambda_{max}$, and it is scale-invariant, i.e., it does not depend on the normalization of $\mathbf{W}$ or $\mathbf{X}$.
The $\lambda_{max}$ is a measure of the size, or scale, of $\mathbf{W}$.
Multiplying each $\alpha$ by the corresponding $\log\lambda_{max}$ weighs ``bigger'' layers more, and averaging this product leads to a balanced, Weighted Alpha metric $\hat{\alpha}$ for the entire~DNN.
We will see that for well-trained CV and NLP models, $\hat{\alpha}$ performs quite well and as expected, but for CV and NLP models that are potentially-problematic or less well-trained, metrics that depend on the scale of the problem can perform anomalously.  
In these cases, separating $\hat{\alpha}$ into its two components, $\alpha$ and $\lambda_{max}$, and examining the distributions of each, can be helpful.

@@

XXX.  THIS IS A VERY PRACTICAL PROBLEM THAT OUR THEORETICAL APPROACH, STRONGLY COUPLED AND MOTIVATED BY EMPIRICAL PRACTICE, CAN SOLVE.

\cite{MM20a_trends_TR} is Charles: ``Predicting trends in the quality of state-of-the-art neural networks without access to training or testing data''

XXX.  ABSTRACT FROM \cite{MM20a_trends_TR}:
In many applications, one works with neural network models trained by someone else.
For such pretrained models, one may not have access to training data or test data.
Moreover, one may not know details about the model, e.g., the specifics of the training data, the loss function, the hyperparameter values, etc.
Given one or many pretrained models, it is a challenge to say anything about the expected performance or quality of the models.
Here, we address this challenge by providing a detailed meta-analysis of hundreds of publicly-available pretrained models.
We examine norm based capacity control metrics as well as power law based metrics from the recently-developed Theory of Heavy-Tailed Self Regularization.
We find that norm based metrics correlate well with reported test accuracies for well-trained models, but that they often cannot distinguish well-trained versus poorly-trained models.
We also find that power law based metrics can do much better---quantitatively better at discriminating among series of well-trained models with a given architecture; and qualitatively better at discriminating well-trained versus poorly-trained models.
These methods can be used to identify when a pretrained neural network has problems that cannot be detected simply by examining training/test accuracies.

XXX.  CITE ALSO BARTLETT AND POGGIO AS DOING SOMETHING LOOSELY SIMILAR, AND THE GOOGLE CONTEST AS DOING SOMETHING LOOSELY SIMILAR BUT DIFFERENT.

\cite{BFT17_TR} is Bartlett: ``Spectrally-normalized margin bounds for neural networks''

\cite{LMBx18_TR} is Poggio: ``A surprising linear relationship predicts test performance in deep networks''

\subsection{Introduction OF PREDICTING TRENDS PAPER}
\label{sxn:intro}

XXX.  THIS IS THE INTRO FROM THE PREDICTING TRENDS PAPER.

A common problem in machine learning (ML) 
is to evaluate the quality of a given model.
A popular way to accomplish this
is to train a model and then evaluate its training/testing error.
There are many problems with this approach.
The training/testing curves give very limited insight into the overall properties of the model; 
they do not take into account the (often large human and CPU/GPU) time for hyperparameter fiddling;
they typically do not correlate with other properties of interest such as robustness or fairness or interpretability; 
and so on.
A related problem, in particular in industrial-scale artificial intelligence (AI), arises when the model user is not the model developer.
Then, one may not have access to the training data or the testing data.
Instead, one may simply be given a model that has already been trained---a pretrained model---and need to use it as-is, or to fine-tune and/or compress it and then use it.

Na\"{\i}vely---but in our experience commonly, among ML practitioners and ML theorists---if one does not have access to training or testing data, then one can say absolutely nothing about the quality of a ML model.
This may be true in worst-case theory, but models are used in practice, and there is a need for a practical theory to guide that practice.
Moreover, if ML is to become an industrial process, then that process will become compartmentalized in order to scale: some groups will gather data, other groups will develop models, and other groups will use those models.
Users of models cannot be expected to know the precise details of how models were built, the specifics of data that were used to train the model, what was the loss function or hyperparameter values, how precisely the model was regularized,~etc.

Moreover, for many large scale, practical applications, there is no obvious way to define an ideal test metric. 
For example, models that generate fake text or conversational chatbots may use a proxy, like perplexity, as a test metric.
In the end, however, they really require human evaluation. 
Alternatively, models that cluster user profiles, which are widely used in areas such as marketing and advertising, are unsupervised and have no obvious labels for comparison and/or evaluation.
In these and other areas, ML objectives can be poor proxies for downstream goals.

>>>>

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Using Heavy-Tailed Self-Regularization Theory}

@@

\twocolumn[
\icmltitle{Using Heavy-Tailed Self-Regularization Theory to Characterize Correlations in and Perform Diagnostics on State-of-the-art Overparameterized Models}

@@

\begin{abstract}
Understanding state-of-the-art overparameterized neural networks requires detailed empirical studies as well as the development of predictive theory in light of those studies. 
Building on the theory of Heavy-Tailed Self-Regularization, we show that correlations in the weight matrices of state-of-the-art overparameterized neural networks are well-described by heavy-tailed random matrix theory.
Using the publicly-available WeightWatcher tool, we perform spectral analylsis on these weight matrices, and we show that fitted truncated power law exponents are strongly predictive of model quality, for publicly-available state-of-the-art pre-trained models, even in the absence of training or testing data.
We also show that these metrics can also be used to identify a Simpson's paradox in the data from a recent contest aimed at developing metrics to predice generalization in deep learning.
\end{abstract}

@@

\michael{Here is Charles core dump of intro; still to refine and incorporate with rest of text.}
We present our Theory of Heavy Tailed Self Regularization (HTSR) and
its application to the (overparameterized)  Deep Neural Networks (DNNs) 
to  predict the generalization accuracy and improve fine-tuning
of pretrained models by capturing the dominant correlation structure.
We analyze the correlations using techniques from 
Heavy Tailed Random Matrix Theory (HT-RMT)
and from  strongly correlated systems from theoretical
chemistry and  physics.  
We show how to predict trends in the reported test accuracies
across hundreds of pretrained DNNs without needing access
to training or test data by charactering the heavy tailed
structure of the empirical spectral densities (ESD) of 
the layer weight matrices.
In doing so, we show that different metrics, which
either depend on the shape or scale of the ESDs,
exhbit a Simpson's paradox when varying DNN model depth
vs regularization hyperparameters.
In particular,  the familar scale-based spectral norm
metric behave opposite theoretical predictions,
whereas our HT alpha-metric acts as predicted.
Moreover, the paradoxical behavior  can be resolved
by combining these shape and scale metrics with
our HT alpha-hat metric.
Finally,  And with access to training data, we show how to
smooth and capture the low rank structure to both accurately
predict the generalization errors and potentially improve
finetuning of pretrained DNNs.

@@

%\citep{MM17_TR} is Charles: Rethinking generalization requires revisiting old ideas: statistical mechanics approaches and complex learning behavior''
%
%\citep{MM18_TR} is Charles: ``Implicit Self-Regularization in Deep Neural Networks: Evidence from Random Matrix Theory and Implications for Learning''
%
%\citep{MM19_HTSR_ICML} is Charles: ``Traditional and Heavy-Tailed Self Regularization in Neural Network Models''
%
%\citep{MM20_SDM} is Charles: ``Heavy-Tailed {U}niversality Predicts Trends in Test Accuracies for Very Large Pre-Trained Deep Neural Networks''
%
%\citep{MM19_KDD} is Charles: ``Statistical Mechanics Methods for Discovering Knowledge from Modern Production Quality Neural Networks''

Recently, Random Matrix Theory (RMT) was applied to analyze empirically the weight matrices of Deep Neural Networks (DNNs), including both production quality, pre-trained models such as AlexNet and Inception, and smaller models trained from scratch, such as LeNet5 and a miniature-AlexNet~\citep{MM18_TR} (see also \citep{MM17_TR, MM18_TR, MM19_HTSR_ICML, MM20_SDM, MM19_KDD}).
%%Empirical results clearly indicate that the empirical spectral density (ESD) of DNN layer matrices displays signatures of traditionally-regularized statistical models, even in the absence of exogenously specifying traditional forms of explicit regularization, such as dropout or weight norm~constraints.
%%
Building on these empirical results, 
%%and using relatively recent results in RMT, most notably its extension to Universality classes of heavy-tailed matrices, 
\citet{MM18_TR} then developed a phenomenological (statistical mechanics based, as opposed to statistical learning theory based) theory to identify ``5+1 Phases of Training,'' corresponding to increasing amounts of implicit self-regularization.
These phases can be observed during the training process as well as in the final learned DNNs.
For smaller and/or older DNNs, this implicit self-regularization is like traditional Tikhonov regularization, in that there is a ``size scale'' separating signal from noise.
For state-of-the-art (SOTA) DNNs, however, they identified a novel form of \emph{Heavy-Tailed Self-Regularization} (HT-ST).
This is similar to the self-organization seen in the statistical physics of disordered systems (such as classical models of actual neural activity); and 
it results from correlations arising at all size scales.
For DNNs, this arises due to the training process itself, when applied to very overparameterized models.

In the HT-SR Theory, one analyzes the eigenvalue spectrum, i.e., the Empirical Spectral Density (ESD), of the associated correlation matrices~\citep{MM18_TR,MM19_HTSR_ICML,MM20_SDM}.
From this, one characterizes the amount and form of correlation present in the DNN's weight matrices.
For each layer weight matrix $\mathbf{W}$, if we compute the eigenvalue spectrum of 
$ \mathbf{X} = \frac{1}{N}\mathbf{W}^{T}\mathbf{W} $
then HT-SR theory characterizes the correlations in a weight matrix by examining its ESD, $\rho(\lambda)$.
Empirically, the ESD can be well-fit to a truncated power law (PL) distribution, given as
$\rho(\lambda)\sim\lambda^{-\alpha} $,
which is (at least) valid within a bounded range of eigenvalues $\lambda\in[\lambda^{min},\lambda^{max}]$.

The original work on HT-SR Theory showed that for nearly every $\mathbf{W}$, i.e., among publicly-available SOTA models, the (bulk and tail) of the ESDs can be fit to a truncated PL, and that PL exponents $\alpha$ nearly all lie within the range $\alpha\in(1.5,5)$~\citep{MM18_TR,MM19_HTSR_ICML,MM20_SDM}.
%
%%As for the mechanism responsible for these properties, statistical physics offers several possibilities~\citep{SornetteBook,nishimori01}, e.g., self organized criticality~\citep{SOC87,SOCat25yrs} or multiplicative noise in the stochastic optimization algorithms used to train these models~\citep{HodMah20A_TR,SorCon97}.
%%Alternatively, related techniques have been used to analyze correlations and information propogation in actual spiking neurons~\citep{SYYRP11,YKYP14}.
%%Our meta-analysis does not require knowledge of mechanisms; and it is not even clear that one mechanism is responsible for every case.
%
Crucially, HT-SR Theory predicts that smaller values of $\alpha$ should correspond to models with better correlation over multiple size scales and thus to better models.
%%The notion of ``size scale'' is well-defined in physical systems, to which this style of analysis is usually applied, but it is less well-defined in CV and NLP applications.
%%Informally, it would correspond to pixel groups that are at a greater distance in some metric, or between sentence parts that are at a greater distance in text.
%%Relatedly, previous work observed that smaller exponents $\alpha$ correspond to more implicit self-regularization and better generalization, and that we expect a linear correlation between $\hat{\alpha}$ and model quality~\citep{MM18_TR,MM19_HTSR_ICML,MM20_SDM}.

\section{Overall approach}

Test accuracies have been reported online for publicly-available pretrained pyTorch models~\citep{osmr}.
These models have been trained and evaluated on labeled data, using standard techniques.  
We do not have access to this data, and we have not trained any of the models ourselves. 
Our methodological approach is thus similar to a statistical meta-analysis, common in biomedical research, but uncommon in machine learning.
%
To evaluate the quality of SOTA DNNs, we consider the following metrics:
Frobenius Norm, $\Vert\mathbf{W}\Vert^{2}_{F}=\Vert\mathbf{X}\Vert_{F}=\sum\nolimits_{i=1}^{M} \lambda_{i}$; 
Spectral Norm, $\Vert\mathbf{W}\Vert_{\infty}^{2}=\Vert\mathbf{X}\Vert_{\infty}=\lambda_{max}$;
Weighted Alpha, $\hat{\alpha}=\alpha\log\lambda_{max}$; and
$\alpha$-Shatten Norm, $\Vert\mathbf{W}\Vert^{2\alpha}_{2\alpha}=\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}=\sum\nolimits_{i=1}^{M}\lambda_{i}^{\alpha}$. 
%\begin{eqnarray*}
%& & \text{Frobenius Norm: $\Vert\mathbf{W}\Vert^{2}_{F}=\Vert\mathbf{X}\Vert_{F}=\sum\nolimits_{i=1}^{M} \lambda_{i}$ } \\
%& & \text{Spectral Norm: $\Vert\mathbf{W}\Vert_{\infty}^{2}=\Vert\mathbf{X}\Vert_{\infty}=\lambda_{max}$ } \\
%& & \text{Weighted Alpha: $\hat{\alpha}=\alpha\log\lambda_{max}$ } \\
%& & \text{$\alpha$-Shatten Norm: $\Vert\mathbf{W}\Vert^{2\alpha}_{2\alpha}=\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}=\sum\nolimits_{i=1}^{M}\lambda_{i}^{\alpha}$. }
%\end{eqnarray*}
To perform diagnostics on potentially-problematic DNNs, we will decompose $\hat{\alpha}$ into its two components, $\alpha$ and $\lambda_{max}$.
One determines $\alpha$ for a given layer by fitting the ESD of that layer's weight matrix to a truncated PL, using the commonly accepted Maximum Likelihood method~\citep{CSN09_powerlaw,ABP14}.
Computations were performed with the publicly-available \texttt{WeightWatcher} tool (version 0.2.7)~\citep{weightwatcher_package}.

@@

%XXX.  ABSTRACT FROM \citep{MM20a_trends_TR}: ``Predicting trends in the quality of state-of-the-art neural networks without access to training or testing data''

In many applications, one works with neural network models trained by someone else.
For such pretrained models, one may not have access to training data or test data.
Moreover, one may not know details about the model, e.g., the specifics of the training data, the loss function, the hyperparameter values, etc.
Given one or many pretrained models, it is a challenge to say anything about the expected performance or quality of the models.
We have applied HT-ST Theory to provide a detailed meta-analysis of hundreds of publicly-available pretrained models, examining both 
norm based capacity control metrics from statistical learning theory as well as PL based metrics from HT-ST theory.

To answer this question, we provide a detailed empirical analysis, evaluating quality metrics for pretrained DNN models, and we do so at scale.
Our approach may be viewed as a statistical meta-analysis of previously published work, where we consider a large suite of hundreds of publicly-available models, mostly from computer vision (CV) and natural language processing (NLP).
By now, there are many such state-of-the-art models that are publicly-available, e.g., 
hundreds of pretrained models in CV ($\ge 500$) and NLP ($\approx 100$).%
\footnote{When we began this work in 2018, there were fewer than tens of such models; now in 2020, there are hundreds of such models; and we expect that in a year or two there will be an order of magnitude or more of such models.}
For all these models, we have no access to training data or testing data, and we have no specific knowledge of the training/testing protocols.

We find that norm based metrics correlate well with reported test accuracies for well-trained models, but that they often cannot distinguish well-trained versus poorly-trained models.
We also find that power law based metrics can do much better---quantitatively better at discriminating among series of well-trained models with a given architecture; and qualitatively better at discriminating well-trained versus poorly-trained models.
These methods can be used to identify when a pretrained neural network has problems that cannot be detected simply by examining training/test accuracies.
%
%Here is a summary of our main results.
%First, norm-based metrics do a reasonably good job at predicting quality trends in well-trained CV/NLP models.
%Second, norm-based metrics may give spurious results when applied to poorly-trained models (e.g., models trained without enough data, etc.).
%For example, they may exhibit what we call Scale Collapse for these models. 
%Third, PL-based metrics can do much better at predicting quality trends in pretrained CV/NLP models.  
%In particular, 
%a weighted PL exponent (weighted by the log of the spectral norm of the corresponding layer) is 
%quantitatively better at discriminating among a series of well-trained versus very-well-trained models
%within a given architecture series; and
%the (unweighted) average PL exponent is 
%qualitatively better at discriminating well-trained versus poorly-trained models.
%Fourth, PL-based metrics can also be used to characterize fine-scale model properties, including what we call layer-wise Correlation Flow, in well-trained and poorly-trained models; and they can be used to evaluate model enhancements (e.g., distillation, fine-tuning, etc.).
%
Our work provides a theoretically-principled empirical evaluation---by far the largest, most detailed, and most comprehensive to date---and it uses HT-SR theory.%
\footnote{Performing such a meta-analysis of previously-published work is common in certain areas, but it is quite rare in ML, where the emphasis is on developing better training protocols.}

%\citep{BFT17_TR} is Bartlett: ``Spectrally-normalized margin bounds for neural networks''
%\citep{LMBx18_TR} is Poggio: ``A surprising linear relationship predicts test performance in deep networks''
There is related work \citep{BFT17_TR,LMBx18_TR}, but that work provides generalization bounds, modifies the models, requires access to data, etc., and so is quite different.
